---
title: "Getting Started with msma"
author: "Atsushi Kawaguchi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with msma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(msma)
```

# Overview

The **msma** package implements sparse and supervised matrix decomposition for
single-block and multiblock multivariable data. The main function is `msma()`.
The analysis is selected by the supplied inputs:

- `X` only: principal component analysis (PCA);
- `X` and `Y`: partial least squares (PLS);
- `Z`: optional external supervision variable;
- positive regularization parameters: sparse estimation.

`X` and `Y` may be matrices or lists of matrices. All blocks must contain the
same observations in the same row order.

# Simulated data

```{r data}
dat <- simdata(n = 40, rho = 0.8, Xps = 5, Yps = 4, seed = 1)
X <- dat$X[[1]]
Y <- dat$Y[[1]]
set.seed(1)
Z <- rbinom(nrow(X), 1, 0.5)

dim(X)
dim(Y)
```

# PCA

A matrix supplied as `X` produces a single-block PCA.

```{r pca}
fit_pca <- msma(X, comp = 2)
fit_pca
summary(fit_pca)
```

The main X-side results are:

- `wbX`: block-level weights or loadings;
- `sbX`: block-level scores;
- `cpevX`: cumulative percentage of explained variance;
- `avX`: adjusted variance attributable to each component.

```{r pca-results}
fit_pca$wbX
head(fit_pca$sbX[[1]])
fit_pca$cpevX
```

```{r pca-plot, fig.show='hold'}
plot(fit_pca, axes = 1, plottype = "bar", las = 2)
plot(fit_pca, v = "score", axes = 1:2, plottype = "scatter")
```

# Sparse PCA

A positive `lambdaX` introduces sparsity into the X-side block weights.

```{r sparse-pca}
fit_spca <- msma(X, comp = 2, lambdaX = 0.10)
fit_spca$nzwbX
fit_spca$selectXnames
```

# Supervised sparse PCA

`Z` supplies external supervision. The strength of supervision on X is
controlled by `muX`.

```{r supervised-pca}
fit_sup_pca <- msma(
  X = X, Z = Z, comp = 2,
  lambdaX = 0.05, muX = 0.20,
  intseed = 1
)
fit_sup_pca$predictiv
```

# PLS

Supplying both `X` and `Y` produces PLS.

```{r pls}
fit_pls <- msma(X = X, Y = Y, comp = 2)
fit_pls
```

```{r pls-plot, fig.show='hold'}
plot(fit_pls, axes = 1, XY = "XY")
plot(fit_pls, axes = 2, XY = "XY")
```

Sparse and supervised PLS are requested by adding `lambdaX`, `lambdaY`, and
optionally `Z`, `muX`, and `muY`.

```{r sparse-supervised-pls}
fit_spls <- msma(
  X = X, Y = Y, Z = Z, comp = 2,
  lambdaX = 0.10, lambdaY = 0.10,
  muX = 0.10, muY = 0.10,
  intseed = 1
)
fit_spls$nzwbX
fit_spls$nzwbY
```

# Prediction

```{r prediction}
pred <- predict(fit_pls, newX = X, newY = Y)
names(pred)
```

# Session information

```{r session-info}
sessionInfo()
```


