---
title: "NMF-Based Super-Level Analysis and ECMSMA"
author: "Atsushi Kawaguchi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{NMF-Based Super-Level Analysis and ECMSMA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(msma)
```

# Overview

Version 3.2 adds NMF and sparse NMF (sNMF) as super-level decomposition
methods. They support clustering-oriented analyses of multiblock data while
the default `sprmethod = "PCA"` preserves the version 3.1 computational path.

```{r data}
dat <- simdata(n = 45, rho = 0.8, Xps = c(5, 5, 5), Yps = 3, seed = 3)
X <- dat$X
names(X) <- paste0("block", seq_along(X))
```

# Super-level methods

The principal options are:

- `sprmethod = "PCA"`: conventional super-level PCA;
- `sprmethod = "NMF"`: non-negative matrix factorization;
- `sprmethod = "sNMF"`: sparse non-negative matrix factorization.

```{r compare-methods}
fit_pca <- msma(X, comp = c(2, 2), sprmethod = "PCA", intseed = 1)
fit_nmf <- msma(X, comp = c(2, 2), sprmethod = "NMF", intseed = 1)
fit_snmf <- msma(
  X, comp = c(2, 2),
  sprmethod = "sNMF",
  lambdaXsup = 0.05,
  intseed = 1
)
```

# Non-negative transformations

NMF requires non-negative input. The `nneg` argument controls transformation
of the block scores:

- `"posneg"`: separates positive and negative parts;
- `"absolute"`: uses absolute values;
- `"min"`: shifts values to a non-negative range.

For a score vector \(s\), positive-negative decomposition is

\[
s^+ = \max(s,0), \qquad s^- = \max(-s,0), \qquad s=s^+-s^-.
\]

```{r nneg-options, eval=FALSE}
fit_posneg <- msma(X, comp = c(2, 2), sprmethod = "NMF", nneg = "posneg")
fit_absolute <- msma(X, comp = c(2, 2), sprmethod = "NMF", nneg = "absolute")
fit_min <- msma(X, comp = c(2, 2), sprmethod = "NMF", nneg = "min")
```

# Non-negativity and reproducibility

```{r nonnegative}
all(unlist(fit_snmf$ssX) >= 0)
all(unlist(fit_snmf$wsX) >= 0)
```

```{r reproducibility}
fit_snmf_2 <- msma(
  X, comp = c(2, 2),
  sprmethod = "sNMF",
  lambdaXsup = 0.05,
  intseed = 1
)
all.equal(fit_snmf$ssX, fit_snmf_2$ssX)
all.equal(fit_snmf$wsX, fit_snmf_2$wsX)
```

# Clustering solutions

A simple cluster assignment is obtained from the largest super-score value for
each observation. Each root component supplies one clustering solution.

```{r clustering}
cluster_matrix <- vapply(
  fit_snmf$ssX,
  function(score) max.col(score, ties.method = "first"),
  integer(nrow(fit_snmf$ssX[[1]]))
)
colnames(cluster_matrix) <- names(fit_snmf$ssX)
head(cluster_matrix)
apply(cluster_matrix, 2, table)
```

The resulting matrix may be supplied to a separate consensus or ensemble
clustering procedure. The consensus step is not performed automatically by
`msma()`.

# Multiple supervision variables

`Z` may be a numeric matrix. `con4spv` specifies weights used to combine its
columns. This is a composite-supervision model rather than a multi-task model
with a separate loading for every outcome.

```{r multivariate-z}
set.seed(3)
z1 <- rnorm(nrow(X[[1]]))
z2 <- 0.5 * z1 + rnorm(nrow(X[[1]]), sd = 0.5)
Z <- cbind(clinical = z1, biomarker = z2)

fit_multi_z <- msma(
  X = X, Z = Z,
  con4spv = c(0.7, 0.3),
  comp = 2,
  muX = 0.20,
  intseed = 1
)
fit_multi_z$predictiv
```

Conceptually, the combined supervision score is

\[
z_c = Zc,
\]

where the supplied weights are normalized internally as required by the
implementation.

# One-column matrix compatibility

A vector and a one-column matrix produce the same result.

```{r one-column-z}
fit_z_vector <- msma(X, Z = z1, comp = 1, muX = 0.20, intseed = 1)
fit_z_matrix <- msma(X, Z = cbind(z1), comp = 1, muX = 0.20, intseed = 1)
fit_z_vector$call <- NULL
fit_z_matrix$call <- NULL
isTRUE(all.equal(fit_z_vector, fit_z_matrix, tolerance = 1e-8))
```

# Practical recommendations

- Fix `intseed` for reproducible NMF and sNMF results.
- Inspect both `ssX` and `wsX` when interpreting clusters.
- Compare several values of `lambdaXsup` for sNMF.
- Treat clustering from different root components as distinct candidate
  solutions before ensemble integration.

# Session information

```{r session-info}
sessionInfo()
```


