---
title: "Multiblock and Nested Component Analysis"
author: "Atsushi Kawaguchi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Multiblock and Nested Component Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(msma)
```

# Multiblock data

Multiblock data are represented by a list of matrices. Rows are observations,
columns are variables, and every block must have the same number of rows.

```{r data}
dat <- simdata(
  n = 40, rho = 0.8,
  Xps = c(4, 5), Yps = c(3, 4),
  seed = 2
)
X <- dat$X
Y <- dat$Y
names(X) <- c("X_block_1", "X_block_2")
names(Y) <- c("Y_block_1", "Y_block_2")
lapply(X, dim)
lapply(Y, dim)
```

# Multiblock PCA

```{r mb-pca}
fit_mb_pca <- msma(X = X, comp = 2, intseed = 1)
fit_mb_pca
```

For X, the fitted object contains block-level weights and scores (`wbX`,
`sbX`) and super-level weights and scores (`wsX`, `ssX`).

```{r mb-components}
lapply(fit_mb_pca$wbX, dim)
lapply(fit_mb_pca$sbX, dim)
lapply(fit_mb_pca$wsX, dim)
lapply(fit_mb_pca$ssX, dim)
```

```{r mb-pca-plot, fig.show='hold'}
plot(fit_mb_pca, axes = 1, plottype = "bar", block = "block", las = 2)
plot(fit_mb_pca, axes = 1, plottype = "bar", block = "super")
```

# Sparse multiblock PCA

`lambdaX` has one value per X block. `lambdaXsup` controls sparsity at the
super level.

```{r sparse-mb-pca}
fit_sparse <- msma(
  X = X, comp = 2,
  lambdaX = c(0.10, 0.15),
  lambdaXsup = 0.05,
  intseed = 1
)
fit_sparse$nzwbX
fit_sparse$nzwsX
```

# Nested components

A two-element `comp` specifies the numbers of root and super components:

```r
comp = c(number_of_root_components, number_of_super_components)
```

For example, `comp = c(2, 3)` estimates three super components for each of two
root components.

```{r nested}
fit_nested <- msma(X = X, comp = c(2, 3), intseed = 1)
lapply(fit_nested$wsX, dim)
lapply(fit_nested$ssX, dim)
```

```{r nested-plot, fig.show='hold'}
plot(fit_nested, axes = 1, axes2 = 1, plottype = "bar", block = "super")
plot(fit_nested, axes = 1, axes2 = 2, plottype = "bar", block = "super")
```

# Supervised multiblock PCA

```{r supervised-mb-pca}
set.seed(2)
Z <- rnorm(nrow(X[[1]]))
fit_supervised <- msma(
  X = X, Z = Z, comp = 2,
  lambdaX = c(0.10, 0.10),
  muX = 0.20,
  intseed = 1
)
fit_supervised$predictiv
```

# Multiblock PLS

```{r mb-pls}
fit_mb_pls <- msma(
  X = X, Y = Y, comp = 2,
  lambdaX = c(0.10, 0.10),
  lambdaY = c(0.10, 0.10),
  intseed = 1
)
fit_mb_pls
```

The four regularization arguments have distinct roles:

- `lambdaX` and `lambdaY`: block-level weights;
- `lambdaXsup` and `lambdaYsup`: super-level weights.

```{r nested-mb-pls}
fit_nested_pls <- msma(
  X = X, Y = Y, comp = c(2, 2),
  lambdaX = c(0.10, 0.10),
  lambdaY = c(0.10, 0.10),
  lambdaXsup = 0.05,
  lambdaYsup = 0.05,
  intseed = 1
)
lapply(fit_nested_pls$ssX, dim)
lapply(fit_nested_pls$ssY, dim)
```

# Session information

```{r session-info}
sessionInfo()
```


