---
title: "Model and Parameter Selection in msma"
author: "Atsushi Kawaguchi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Model and Parameter Selection in msma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(msma)
```

# Overview

The package provides functions for selecting component numbers and
regularization parameters:

- `ncompsearch()`: number of components;
- `regparasearch()`: regularization parameters;
- `optparasearch()`: combined selection.

BIC is useful for a quick deterministic search. Cross-validation may be more
computationally expensive.

```{r data}
dat <- simdata(n = 35, rho = 0.8, Xps = c(4, 4), Yps = c(3, 3), seed = 4)
X <- dat$X
Y <- dat$Y
```

# Selecting the number of components

```{r ncomp-bic}
search_comp <- ncompsearch(X, comps = 1:3, criterion = "BIC", intseed = 1)
search_comp
```

```{r ncomp-plot}
plot(search_comp)
```

For nested analysis, candidates may be supplied as a list for root and super
components.

```{r nested-search, eval=FALSE}
search_nested <- ncompsearch(
  X,
  comps = list(1:4, 1:3),
  criterion = "BIC",
  intseed = 1
)
```

# Selecting regularization parameters

The following example is shown but not evaluated during package building to
keep the vignette lightweight.

```{r regularization-search, eval=FALSE}
search_lambda <- regparasearch(
  X = X,
  comp = 2,
  criterion = "BIC",
  maxrep = 5,
  intseed = 1
)
search_lambda
```

# Combined selection

`optparasearch()` supports four workflows:

- `"regparaonly"`: regularization search for fixed components;
- `"ncomp1st"`: component search followed by regularization search;
- `"regpara1st"`: regularization search followed by component search;
- `"simultaneous"`: repeated joint search.

```{r combined-search, eval=FALSE}
opt <- optparasearch(
  X = X,
  search.method = "ncomp1st",
  criterion = "BIC",
  intseed = 1
)

fit <- msma(
  X = X,
  comp = opt$optncomp,
  lambdaX = opt$optlambdaX,
  lambdaXsup = opt$optlambdaXsup,
  intseed = 1
)
```

# PLS selection

X-side and Y-side parameters are selected separately in PLS.

```{r pls-selection, eval=FALSE}
opt_pls <- optparasearch(
  X = X, Y = Y,
  search.method = "regparaonly",
  criterion = "BIC",
  intseed = 1
)

fit_pls <- msma(
  X = X, Y = Y,
  comp = opt_pls$optncomp,
  lambdaX = opt_pls$optlambdaX,
  lambdaY = opt_pls$optlambdaY,
  lambdaXsup = opt_pls$optlambdaXsup,
  lambdaYsup = opt_pls$optlambdaYsup,
  intseed = 1
)
```

# Cross-validation

```{r cv-example, eval=FALSE}
cv <- cvmsma(
  X = X, Y = Y,
  comp = 1,
  lambdaX = c(0.1, 0.1),
  lambdaY = c(0.1, 0.1),
  nfold = 5,
  seed = 1,
  intseed = 1
)
cv
```

# Version 3.2 super-level methods

Model-selection functions accept the super-level method arguments where
applicable.

```{r snmf-selection, eval=FALSE}
search_snmf <- ncompsearch(
  X,
  comps = list(1:3, 1:3),
  criterion = "BIC",
  sprmethod = "sNMF",
  nneg = "posneg",
  intseed = 1
)
```

# Computational recommendations

- Start with a small candidate grid.
- Use BIC to screen candidate models before cross-validation.
- Fix `seed` for fold allocation and `intseed` for model estimation.
- Keep expensive searches as explicit user actions rather than vignette build
  steps.

# Session information

```{r session-info}
sessionInfo()
```


