---
title: "msma 3.2: Version 3.1 Compatibility Test"
author: "Atsushi Kawaguchi"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  legacy_source: "msma_3.1.tar.gz"
  latest_source: "../src.r"
  tolerance: 1.0e-8
  strict: true
  run_cv: true
  run_search: false
---

# Purpose

This document checks whether **msma 3.2** reproduces the results of the
existing **msma 3.1** implementation when only the pre-existing/default
functionality is used.

The new features introduced in version 3.2, including
`sprmethod = "sNMF"` and matrix-valued `Z`, are intentionally excluded from
the backward-compatibility comparisons. Separate smoke tests for these new
features are included at the end of this document.

The comparison is performed by loading the two implementations into separate
R environments. Therefore, the functions with the same names do not overwrite
each other.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  collapse = TRUE,
  comment = "#>"
)

required_packages <- c("knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the following package(s): ", paste(missing_packages, collapse = ", "))
}
```

```{r configuration}
cat("Legacy source:", params$legacy_source, "\n")
cat("Latest source:", params$latest_source, "\n")
cat("Tolerance:", format(params$tolerance, scientific = TRUE), "\n")
cat("Strict mode:", params$strict, "\n")
cat("R version:", R.version.string, "\n")
cat("Platform:", R.version$platform, "\n")
```

# Load the two implementations

The legacy input may be either:

- the CRAN source archive `msma_3.1.tar.gz`;
- an extracted package directory containing `R/src.r`; or
- a standalone R source file.

The latest input is normally the version 3.2 `src.r` file, but an extracted
package directory or source archive is also accepted.

```{r loader-functions}
find_r_sources <- function(path, label) {
  if (!file.exists(path) && !dir.exists(path)) {
    stop(label, " was not found: ", path)
  }

  if (dir.exists(path)) {
    r_dir <- file.path(path, "R")
    if (!dir.exists(r_dir)) {
      stop(label, " is a directory but does not contain an R/ directory: ", path)
    }
    files <- list.files(r_dir, pattern = "[.]r$|[.]R$", full.names = TRUE)
    if (!length(files)) stop("No R source files were found in ", r_dir)
    return(sort(files))
  }

  if (grepl("[.]tar[.]gz$|[.]tgz$", path, ignore.case = TRUE)) {
    exdir <- tempfile(paste0(label, "_"))
    dir.create(exdir)
    utils::untar(path, exdir = exdir)
    candidates <- list.files(
      exdir,
      pattern = "[.]r$|[.]R$",
      recursive = TRUE,
      full.names = TRUE
    )
    candidates <- candidates[grepl("[/\\\\]R[/\\\\]", candidates)]
    if (!length(candidates)) stop("No package R source files were found in ", path)
    return(sort(candidates))
  }

  path
}

load_implementation <- function(path, label) {
  files <- find_r_sources(path, label)
  env <- new.env(parent = globalenv())
  for (f in files) {
    sys.source(f, envir = env, keep.source = TRUE)
  }

  required <- c(
    "msma.default", "predict.msma", "cvmsma",
    "ncompsearch", "regparasearch", "simdata"
  )
  absent <- required[!vapply(required, exists, logical(1), envir = env, inherits = FALSE)]
  if (length(absent)) {
    stop(label, " is missing function(s): ", paste(absent, collapse = ", "))
  }

  list(env = env, files = files)
}

legacy <- load_implementation(params$legacy_source, "legacy_source")
latest <- load_implementation(params$latest_source, "latest_source")

cat("Legacy files loaded:\n", paste(legacy$files, collapse = "\n"), "\n\n")
cat("Latest files loaded:\n", paste(latest$files, collapse = "\n"), "\n")
```

# Comparison utilities

Numerical quantities are compared using `all.equal()` with the tolerance in
`params$tolerance`. Calls and environments are removed before comparison,
because they necessarily refer to different source environments.

```{r comparison-functions}
normalize_for_comparison <- function(x) {
  if (inherits(x, "msma")) {
    x$call <- NULL
  }
  if (is.list(x)) {
    x <- lapply(x, normalize_for_comparison)
    attributes(x) <- attributes(x)
  }
  x
}

compare_objects <- function(old, new, tolerance = params$tolerance) {
  old <- normalize_for_comparison(old)
  new <- normalize_for_comparison(new)
  result <- all.equal(
    old,
    new,
    tolerance = tolerance,
    check.attributes = TRUE
  )
  list(
    equal = isTRUE(result),
    detail = if (isTRUE(result)) "OK" else paste(result, collapse = "; ")
  )
}

results <- list()
add_result <- function(test, object, old, new) {
  z <- compare_objects(old, new)
  results[[length(results) + 1L]] <<- data.frame(
    test = test,
    object = object,
    equal = z$equal,
    detail = z$detail,
    stringsAsFactors = FALSE
  )
  invisible(z)
}

compare_fit_elements <- function(test, old_fit, new_fit, elements) {
  for (nm in elements) {
    old_value <- if (nm %in% names(old_fit)) old_fit[[nm]] else structure(NULL, missing = TRUE)
    new_value <- if (nm %in% names(new_fit)) new_fit[[nm]] else structure(NULL, missing = TRUE)
    add_result(test, nm, old_value, new_value)
  }
  invisible(NULL)
}

run_fit_pair <- function(args) {
  old_args <- args
  new_args <- args
  old_args$X <- args$X
  new_args$X <- args$X

  set.seed(1001)
  old_fit <- do.call(legacy$env$msma.default, old_args)
  set.seed(1001)
  new_fit <- do.call(latest$env$msma.default, new_args)
  list(old = old_fit, new = new_fit)
}
```

# Fixed test data

The same matrices are supplied directly to both implementations. This avoids
confounding compatibility of the fitting functions with compatibility of the
simulation function.

```{r test-data}
set.seed(20250808)
n <- 36L

latent1 <- rnorm(n)
latent2 <- rnorm(n)

X1 <- cbind(
  x11 = latent1 + rnorm(n, sd = 0.25),
  x12 = -0.7 * latent1 + rnorm(n, sd = 0.30),
  x13 = latent2 + rnorm(n, sd = 0.25),
  x14 = rnorm(n),
  x15 = 0.4 * latent1 + 0.3 * latent2 + rnorm(n, sd = 0.35)
)

X2 <- cbind(
  x21 = 0.8 * latent1 + rnorm(n, sd = 0.30),
  x22 = latent2 + rnorm(n, sd = 0.25),
  x23 = -0.6 * latent2 + rnorm(n, sd = 0.30),
  x24 = rnorm(n)
)

Y1 <- cbind(
  y11 = 0.9 * latent1 + rnorm(n, sd = 0.30),
  y12 = latent2 + rnorm(n, sd = 0.30),
  y13 = 0.4 * latent1 - 0.5 * latent2 + rnorm(n, sd = 0.35)
)

Y2 <- cbind(
  y21 = -0.7 * latent1 + rnorm(n, sd = 0.30),
  y22 = 0.8 * latent2 + rnorm(n, sd = 0.25)
)

X <- list(modality1 = X1, modality2 = X2)
Y <- list(outcome1 = Y1, outcome2 = Y2)
Z <- as.numeric(scale(0.8 * latent1 - 0.4 * latent2 + rnorm(n, sd = 0.35)))

str(list(X = X, Y = Y, Z = Z), max.level = 2)
```

# Compatibility tests

## Single-block PCA

```{r test-single-pca}
fit <- run_fit_pair(list(
  X = X1,
  comp = 2,
  intseed = 11
))

compare_fit_elements(
  "Single-block PCA",
  fit$old,
  fit$new,
  c("dmode", "comp", "X", "Xscale", "wbX", "sbX", "ssX", "wsX",
    "nzwbX", "nzwsX", "selectXnames", "cpevX", "avX", "bic", "bic2")
)
```

## Multiblock PCA

```{r test-multiblock-pca}
fit_mb_pca <- run_fit_pair(list(
  X = X,
  comp = c(2, 1),
  intseed = 11
))

compare_fit_elements(
  "Multiblock PCA",
  fit_mb_pca$old,
  fit_mb_pca$new,
  c("dmode", "comp", "X", "Xscale", "wbX", "sbX", "ssX", "wsX",
    "nzwbX", "nzwsX", "selectXnames", "cpevX", "avX", "bic", "bic2")
)
```

## Sparse multiblock PCA

```{r test-sparse-pca}
fit_sparse_pca <- run_fit_pair(list(
  X = X,
  comp = c(2, 1),
  lambdaX = c(0.15, 0.10),
  lambdaXsup = 0.05,
  intseed = 11
))

compare_fit_elements(
  "Sparse multiblock PCA",
  fit_sparse_pca$old,
  fit_sparse_pca$new,
  c("wbX", "sbX", "ssX", "wsX", "nzwbX", "nzwsX", "selectXnames",
    "cpevX", "avX", "bic", "bic2")
)
```

## Supervised multiblock PCA with vector-valued Z

```{r test-supervised-pca}
fit_sup_pca <- run_fit_pair(list(
  X = X,
  Z = Z,
  comp = c(2, 1),
  lambdaX = c(0.10, 0.10),
  muX = 0.20,
  intseed = 11
))

compare_fit_elements(
  "Supervised multiblock PCA",
  fit_sup_pca$old,
  fit_sup_pca$new,
  c("Z", "muX", "wbX", "sbX", "ssX", "wsX", "nzwbX", "nzwsX",
    "cpevX", "avX", "bic", "bic2", "predictiv")
)
```

## Multiblock PLS

```{r test-multiblock-pls}
fit_pls <- run_fit_pair(list(
  X = X,
  Y = Y,
  comp = c(2, 1),
  lambdaX = c(0.10, 0.10),
  lambdaY = c(0.10, 0.10),
  intseed = 11
))

compare_fit_elements(
  "Multiblock PLS",
  fit_pls$old,
  fit_pls$new,
  c("dmode", "comp", "X", "Y", "Xscale", "Yscale", "wbX", "sbX",
    "wbY", "sbY", "ssX", "wsX", "ssY", "wsY", "nzwbX", "nzwbY",
    "nzwsX", "nzwsY", "selectXnames", "selectYnames", "cpevX", "cpevY",
    "avX", "avY", "bic", "bicX", "bicY", "bic2", "bic2X", "bic2Y")
)
```

## Prediction

```{r test-prediction}
set.seed(1001)
pred_old <- legacy$env$predict.msma(fit_pls$old, newX = X, newY = Y)
set.seed(1001)
pred_new <- latest$env$predict.msma(fit_pls$new, newX = X, newY = Y)

for (nm in c("X", "sbX", "Y", "sbY")) {
  add_result("Prediction", nm, pred_old[[nm]], pred_new[[nm]])
}
```

## Cross-validation

```{r test-cv, eval=params$run_cv}
cv_args <- list(
  X = X,
  Y = Y,
  comp = 1,
  lambdaX = c(0.10, 0.10),
  lambdaY = c(0.10, 0.10),
  nfold = 3,
  seed = 71,
  intseed = 11
)

set.seed(1001)
cv_old <- do.call(legacy$env$cvmsma, cv_args)
set.seed(1001)
cv_new <- do.call(latest$env$cvmsma, cv_args)
add_result("Cross-validation", "cvmsma result", cv_old, cv_new)
```

## Optional component search

This test is disabled by default because it takes longer. Set
`params$run_search: true` when rendering to activate it.

```{r test-search, eval=params$run_search}
search_args <- list(
  X = X,
  Y = Y,
  comps = 1:2,
  lambdaX = c(0.10, 0.10),
  lambdaY = c(0.10, 0.10),
  criterion = "BIC",
  intseed = 11
)

set.seed(1001)
search_old <- do.call(legacy$env$ncompsearch, search_args)
set.seed(1001)
search_new <- do.call(latest$env$ncompsearch, search_args)
add_result("Component search", "ncompsearch result", search_old, search_new)
```

# Simulation-function compatibility

```{r test-simulation}
set.seed(1001)
sim_old <- legacy$env$simdata(n = 20, rho = 0.6, Xps = c(4, 3), Yps = c(3, 2), seed = 41)
set.seed(1001)
sim_new <- latest$env$simdata(n = 20, rho = 0.6, Xps = c(4, 3), Yps = c(3, 2), seed = 41)
add_result("Simulation", "simdata result", sim_old, sim_new)
```

# Results

```{r compatibility-results}
result_table <- do.call(rbind, results)
rownames(result_table) <- NULL

knitr::kable(
  result_table,
  caption = "Compatibility results for msma 3.1 and msma 3.2"
)

n_pass <- sum(result_table$equal)
n_fail <- sum(!result_table$equal)
cat("\nPassed:", n_pass, "\n")
cat("Failed:", n_fail, "\n")
```

```{r strict-decision}
if (params$strict && any(!result_table$equal)) {
  failed <- result_table[!result_table$equal, , drop = FALSE]
  stop(
    "Backward-compatibility test failed for ", nrow(failed),
    " comparison(s): ",
    paste(paste(failed$test, failed$object, sep = " / "), collapse = "; ")
  )
}
```

# Version 3.2 new-feature smoke tests

These checks establish that the new functionality runs. They are not expected
to match version 3.1 because the functionality was not available there.

## One-column matrix Z agrees with vector Z

```{r smoke-matrix-z}
set.seed(1001)
new_z_vector <- latest$env$msma.default(
  X = X,
  Z = Z,
  comp = 1,
  muX = 0.20,
  intseed = 11
)

set.seed(1001)
new_z_matrix <- latest$env$msma.default(
  X = X,
  Z = cbind(Z),
  comp = 1,
  muX = 0.20,
  intseed = 11
)

z_check <- compare_objects(new_z_vector, new_z_matrix)
cat("Vector Z versus one-column matrix Z:", z_check$detail, "\n")
if (params$strict && !z_check$equal) stop("Vector/matrix Z equivalence test failed")
```

## Multiple-column Z

```{r smoke-multiple-z}
Z_multi <- cbind(
  clinical = Z,
  biomarker = as.numeric(scale(latent2 + rnorm(n, sd = 0.25)))
)

set.seed(1001)
fit_multi_z <- latest$env$msma.default(
  X = X,
  Z = Z_multi,
  con4spv = c(0.7, 0.3),
  comp = 1,
  muX = 0.20,
  intseed = 11
)

stopifnot(inherits(fit_multi_z, "msma"))
cat("Multiple-column Z test: OK\n")
```

## Sparse NMF

```{r smoke-snmf}
set.seed(1001)
fit_snmf <- latest$env$msma.default(
  X = X,
  comp = c(2, 2),
  sprmethod = "sNMF",
  lambdaXsup = 0.05,
  intseed = 11
)

stopifnot(all(unlist(fit_snmf$ssX) >= 0))
stopifnot(all(unlist(fit_snmf$wsX) >= 0))
cat("sNMF non-negativity test: OK\n")
```

# Session information

```{r session-info}
sessionInfo()
```


